---
title: "Turkish CPI Forecasting Project Using TÜİK Data"
author: "Faysal Yaman - 138721021"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
# Global chunk options: show code, suppress warnings and messages
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 5)
```

# Project Information

| Field | Detail |
|---|---|
| **Student Name** | Faysal Yaman |
| **Student Number** | 138721021 |
| **Data Source** | TÜİK Data Portal via tuikr package |
| **Theme** | Price Statistics (Theme ID: 6) |
| **Table** | Index numbers and rate of changes in the consumer price index |
| **Variable** | CPI General Index (Base year: 2003=100) |
| **Frequency** | Monthly |
| **Period** | January 2005 - December 2025 |
| **Forecast Target** | January 2026 |

# Required Packages

```{r packages}
# Automatically install missing packages
required_packages <- c("httr", "readxl", "dplyr", "tidyr",
                       "lubridate", "ggplot2", "forecast", "zoo", "rmarkdown")
new_packages <- required_packages[!sapply(required_packages,
                                          requireNamespace, quietly = TRUE)]
if (length(new_packages) > 0) install.packages(new_packages)

# Install tuikr from GitHub if not available
if (!requireNamespace("tuikr", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
  devtools::install_github("emraher/tuikr")
}

# Load all required libraries
library(tuikr)     # TÜİK data access
library(httr)      # HTTP requests
library(readxl)    # Excel file reading
library(dplyr)     # Data manipulation
library(tidyr)     # Data reshaping
library(lubridate) # Date operations
library(ggplot2)   # Visualization
library(forecast)  # Time series forecasting
library(zoo)       # Time series utilities
```

# Data Access via tuikr

Data is accessed entirely through the tuikr R package combined with httr::GET().
No manual downloads were performed. This approach was approved by the instructor
because the TÜİK SDMX API (nsiws.tuik.gov.tr) returns HTTP 401 for statistical_data().

```{r data-import}
# Step 1: Use tuikr to discover Price Statistics tables (Theme ID: 6)
price_tables <- statistical_tables("6")

# Step 2: Filter for the CPI istab table
cpi_row <- price_tables |>
  filter(node_type == "istab",
         grepl("Index numbers and rate of changes in the consumer price index",
               table_name, ignore.case = TRUE)) |>
  slice(1)

cat("Table name:", cpi_row$table_name, "
")
cat("Node type :", cpi_row$node_type, "
")

# Step 3: Fetch data programmatically using httr::GET() inside R
cpi_url <- cpi_row$table_url

response <- GET(
  cpi_url,
  add_headers(
    `User-Agent` = "Mozilla/5.0 (Macintosh; Intel Mac OS X 10_15_7) AppleWebKit/537.36",
    `Accept`     = "application/vnd.ms-excel, application/octet-stream, */*",
    `Referer`    = "https://veriportali.tuik.gov.tr/"
  )
)

cat("HTTP Status:", status_code(response), "
")

# Step 4: Save to temporary file and read with readxl
tmp_file <- tempfile(fileext = ".xls")
writeBin(content(response, "raw"), tmp_file)
cpi_raw <- read_excel(tmp_file, col_names = FALSE)

cat("Raw data dimensions:", nrow(cpi_raw), "rows x", ncol(cpi_raw), "columns
")
```

# Data Cleaning and Preparation

```{r data-cleaning}
# The Excel file contains 5 stacked tables.
# Rows 5-25 contain the CPI Index Level table (2003=100 base).
# Rows 27-51: Monthly change %, Rows 53-77: Cumulative change %
# Rows 79-103: Annual change %, Rows 105-129: 12-month moving average

# Extract only the index level table (rows 5 to 25)
cpi_index <- cpi_raw[5:25, ]

# Assign column names: year + 12 months
colnames(cpi_index) <- c("year", "Jan", "Feb", "Mar", "Apr", "May",
                          "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Convert all columns to numeric
cpi_index <- cpi_index |> mutate(across(everything(), as.numeric))

# Reshape from wide format to long format
cpi_long <- cpi_index |>
  pivot_longer(cols = Jan:Dec, names_to = "month", values_to = "index") |>
  mutate(
    month_num = match(month, c("Jan","Feb","Mar","Apr","May","Jun",
                               "Jul","Aug","Sep","Oct","Nov","Dec")),
    date = as.Date(paste(year, month_num, "1", sep = "-"))
  ) |>
  filter(!is.na(index)) |>
  arrange(date) |>
  select(date, index)

# Create time series object (monthly frequency starting January 2005)
cpi_ts <- ts(cpi_long$index, start = c(2005, 1), frequency = 12)

# Summary of the cleaned data
cat("=== DATA SUMMARY ===
")
cat("Total observations :", nrow(cpi_long), "
")
cat("First observation  :", as.character(min(cpi_long$date)), "
")
cat("Last observation   :", as.character(max(cpi_long$date)), "
")
cat("Forecast target    : January 2026
")
cat("Missing values     :", sum(is.na(cpi_long$index)), "
")
cat("Minimum index      :", round(min(cpi_long$index), 2), "
")
cat("Maximum index      :", round(max(cpi_long$index), 2), "
")
cat("Mean index         :", round(mean(cpi_long$index), 2), "
")

# Display first and last observations
head(cpi_long, 6)
tail(cpi_long, 6)
```

# Exploratory Time Series Analysis

## Actual Series Plot

```{r eda-plot}
# Plot the full CPI time series
p_actual <- ggplot(cpi_long, aes(x = date, y = index)) +
  geom_line(color = "#2C3E50", linewidth = 0.8) +
  geom_point(data = tail(cpi_long, 1), aes(x = date, y = index),
             color = "red", size = 3) +
  labs(
    title    = "Turkish CPI General Index (2003=100) - Monthly",
    subtitle = "January 2005 - December 2025",
    x        = "Date",
    y        = "Index Value",
    caption  = "Source: TÜİK, accessed via tuikr package"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

print(p_actual)
ggsave("outputs/figures/actual_series_plot.png", p_actual,
       width = 10, height = 5, dpi = 150)
```

## Decomposition Analysis

```{r decomp-plot}
# Multiplicative decomposition: appropriate when seasonal variation grows with trend
decomp <- decompose(cpi_ts, type = "multiplicative")

png("outputs/figures/decomposition_plot.png", width = 1000, height = 700)
plot(decomp, col = "#2C3E50")
dev.off()

plot(decomp, col = "#2C3E50")
```

## Exploratory Analysis Summary

**Trend:** The series shows a very strong and persistent upward trend. The CPI index
increased from approximately 114 in January 2005 to 3514 in December 2025, representing
a more than 30-fold increase over 20 years.

**Seasonality:** A recurring seasonal pattern is present. Index values tend to
increase more in January-March (post-holiday price adjustments) and in
autumn months (September-October).

**Structural Break:** A significant acceleration in the rate of increase is visible
from 2022 onwards, corresponding to the Turkish currency crisis and subsequent
high inflation period. This structural break makes some classical methods less effective.

**Missing Values:** None detected. The series is complete with 252 monthly observations.

**Outliers:** The 2022-2023 period shows unusually large monthly increases but these
reflect real economic conditions rather than data errors.

# Accuracy Measure Functions

```{r accuracy-fn}
# Function to compute all 7 required accuracy measures
calc_accuracy <- function(actual, fitted, method_name) {
  errors  <- actual - fitted                                    # Forecast errors
  bias    <- mean(errors, na.rm = TRUE)                        # Bias / Mean Error
  mad     <- mean(abs(errors), na.rm = TRUE)                   # MAD
  mse     <- mean(errors^2, na.rm = TRUE)                      # MSE
  mape    <- mean(abs(errors / actual) * 100, na.rm = TRUE)    # MAPE
  rsfe    <- sum(errors, na.rm = TRUE)                         # RSFE
  ts_val  <- rsfe / mad                                        # Tracking Signal

  data.frame(
    Method          = method_name,
    Bias_ME         = round(bias,   3),
    MAD             = round(mad,    3),
    MSE             = round(mse,    3),
    MAPE            = round(mape,   3),
    RSFE            = round(rsfe,   3),
    Tracking_Signal = round(ts_val, 3)
  )
}

# Helper function to create actual vs forecast plots
plot_forecast <- function(actual_df, fitted_vals, forecast_val,
                          forecast_date, method_name, filename) {
  fitted_df   <- data.frame(date = actual_df$date, value = fitted_vals)
  forecast_df <- data.frame(date = forecast_date,  value = forecast_val)

  p <- ggplot(actual_df, aes(x = date, y = index)) +
    geom_line(color = "#2C3E50", linewidth = 0.7) +
    geom_line(data = fitted_df, aes(x = date, y = value),
              color = "#E74C3C", linewidth = 0.6, linetype = "dashed") +
    geom_point(data = forecast_df, aes(x = date, y = value),
               color = "#27AE60", size = 4, shape = 18) +
    labs(
      title    = paste(method_name, "- Actual vs Fitted + Forecast"),
      subtitle = paste("Forecast (January 2026):", round(forecast_val, 2)),
      x        = "Date",
      y        = "CPI Index (2003=100)",
      caption  = "Black: Actual | Red dashed: Fitted | Green diamond: Forecast"
    ) +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold"))

  print(p)
  ggsave(paste0("outputs/figures/", filename), p,
         width = 10, height = 5, dpi = 150)
}

# Common variables used across all methods
forecast_date <- as.Date("2026-01-01")
accuracy_list <- list()
n      <- nrow(cpi_long)
actual <- cpi_long$index
```

# Forecasting Methods

## Method 1: Naive Forecasting

```{r m1-naive}
# Naive method: each forecast equals the previous actual value
# Formula: Y_hat(t) = Y(t-1)
# This serves as a benchmark for comparing other methods.

naive_fitted   <- c(NA, actual[-n])   # shift actual by 1 period
naive_forecast <- actual[n]           # last observed value is the forecast

cat("Naive Forecast (January 2026):", round(naive_forecast, 2), "
")

# Compute accuracy measures (exclude first NA)
acc1 <- calc_accuracy(actual[-1], naive_fitted[-1], "1. Naive")
accuracy_list[[1]] <- acc1
print(acc1)

# Plot
plot_forecast(cpi_long, naive_fitted, naive_forecast,
              forecast_date, "Naive", "naive_forecast_plot.png")
```

**Interpretation:** The naive method simply carries forward the last observed value
(December 2025: 3513.87) as the forecast for January 2026. With MAPE of 1.437%,
it performs reasonably well for a series with strong trend, but it consistently
underestimates future values because it ignores the upward trend entirely.

## Method 2: Moving Average (k=12)

```{r m2-ma}
# Moving average with window k=12 months.
# A 12-month window is chosen because it covers one full seasonal cycle,
# which helps smooth out seasonal fluctuations while tracking the trend.

k          <- 12
ma_fitted  <- as.numeric(stats::filter(actual, rep(1/k, k), sides = 1))
ma_forecast <- mean(tail(actual, k))   # average of last 12 observations

cat("Moving Average Forecast (January 2026):", round(ma_forecast, 2), "
")

# Accuracy (valid indices only)
vi2  <- which(!is.na(ma_fitted))
acc2 <- calc_accuracy(actual[vi2], ma_fitted[vi2], "2. Moving Average (k=12)")
accuracy_list[[2]] <- acc2
print(acc2)

plot_forecast(cpi_long, ma_fitted, ma_forecast,
              forecast_date, "Moving Average (k=12)", "moving_average_plot.png")
```

**Interpretation:** The 12-month moving average smooths out short-term fluctuations
but significantly lags behind the actual series due to the strong upward trend.
MAPE of 6.934% is the worst among trend-aware methods. The method consistently
underestimates during rapid inflation periods, confirming it is not well-suited
for strongly trending series.

## Method 3: Weighted Moving Average

```{r m3-wma}
# Weighted moving average with 3-period window.
# Weights: 0.5 (most recent), 0.3 (one period ago), 0.2 (two periods ago).
# Higher weights assigned to recent observations to capture recent trend better.

weights      <- c(0.2, 0.3, 0.5)   # oldest to most recent
w            <- length(weights)
wma_fitted   <- rep(NA, n)
for (i in (w+1):n) {
  wma_fitted[i] <- sum(weights * actual[(i-w):(i-1)])
}
wma_forecast <- sum(weights * actual[(n-w+1):n])

cat("Weighted MA Forecast (January 2026):", round(wma_forecast, 2), "
")

vi3  <- which(!is.na(wma_fitted))
acc3 <- calc_accuracy(actual[vi3], wma_fitted[vi3],
                      "3. Weighted MA (0.5/0.3/0.2)")
accuracy_list[[3]] <- acc3
print(acc3)

plot_forecast(cpi_long, wma_fitted, wma_forecast,
              forecast_date, "Weighted Moving Average", "weighted_moving_average_plot.png")
```

**Interpretation:** By assigning higher weight to recent observations, the weighted
MA responds faster to recent changes than the simple MA. MAPE of 2.311% is better
than simple MA but still higher than exponential smoothing methods. The method
works better during stable periods but still struggles with structural breaks.

## Method 4: Simple Exponential Smoothing

```{r m4-ses}
# Simple exponential smoothing (SES): assigns exponentially decreasing weights
# to past observations. Alpha parameter is optimized to minimize MSE.
# Best suited for series without trend or seasonality - expected to underperform here.

ses_model    <- ses(cpi_ts, h = 1)
ses_alpha    <- ses_model$model$par["alpha"]
ses_fitted   <- as.numeric(ses_model$fitted)
ses_forecast <- as.numeric(ses_model$mean[1])

cat("SES Alpha (optimized):", round(ses_alpha, 4), "
")
cat("SES Forecast (January 2026):", round(ses_forecast, 2), "
")

acc4 <- calc_accuracy(actual, ses_fitted, "4. Simple Exp. Smoothing")
accuracy_list[[4]] <- acc4
print(acc4)

plot_forecast(cpi_long, ses_fitted, ses_forecast,
              forecast_date, "Simple Exponential Smoothing", "exponential_smoothing_plot.png")
```

**Interpretation:** SES with optimized alpha produces results very similar to the
naive method (MAPE: 1.433%). The high alpha value means the model heavily weights
the most recent observation, essentially mimicking a naive forecast. SES is not
designed for trending series, so it systematically underestimates future values.

## Method 5: Holt Trend-Corrected Exponential Smoothing

```{r m5-holt}
# Holt double exponential smoothing: extends SES by adding a trend component.
# Two parameters: alpha (level) and beta (trend), both optimized automatically.
# This method is specifically designed for series with trend - well suited here.

holt_model    <- holt(cpi_ts, h = 1)
holt_alpha    <- holt_model$model$par["alpha"]
holt_beta     <- holt_model$model$par["beta"]
holt_fitted   <- as.numeric(holt_model$fitted)
holt_forecast <- as.numeric(holt_model$mean[1])

cat("Holt Alpha:", round(holt_alpha, 4), "| Beta:", round(holt_beta, 4), "
")
cat("Holt Forecast (January 2026):", round(holt_forecast, 2), "
")

acc5 <- calc_accuracy(actual, holt_fitted, "5. Holt Trend-Corrected ES")
accuracy_list[[5]] <- acc5
print(acc5)

plot_forecast(cpi_long, holt_fitted, holt_forecast,
              forecast_date, "Holt Trend-Corrected ES", "trend_adjusted_smoothing_plot.png")
```

**Interpretation:** Holt method achieves the best accuracy with MAPE of only 0.825%.
By explicitly modeling both level and trend components, it captures the upward
trajectory of Turkish CPI much better than simpler methods. The optimized beta
parameter allows the trend estimate to adapt to recent changes in the inflation rate.
This is the superior method for this dataset.

## Method 6: Linear Trend Projection

```{r m6-linear}
# Linear trend projection fits OLS regression: Y = a + b*t
# where t is a time index (1, 2, ..., n).
# Intercept (a): estimated index value at t=0
# Slope (b): average monthly increase in index points

t_var       <- 1:n
lm_model    <- lm(actual ~ t_var)
lm_a        <- coef(lm_model)[1]
lm_b        <- coef(lm_model)[2]
lm_fitted   <- as.numeric(fitted(lm_model))
lm_forecast <- lm_a + lm_b * (n + 1)

cat("Linear Trend Equation: Y =", round(lm_a, 2), "+", round(lm_b, 4), "* t
")
cat("Intercept interpretation: estimated index at start is", round(lm_a, 2), "
")
cat("Slope interpretation: average monthly increase of", round(lm_b, 2), "index points
")
cat("Linear Trend Forecast (January 2026):", round(lm_forecast, 2), "
")

acc6 <- calc_accuracy(actual, lm_fitted, "6. Linear Trend Projection")
accuracy_list[[6]] <- acc6
print(acc6)

plot_forecast(cpi_long, lm_fitted, lm_forecast,
              forecast_date, "Linear Trend Projection", "trend_projection_plot.png")
```

**Interpretation:** The linear trend model fits poorly (MAPE: 118.853%) because
CPI growth is exponential, not linear. The model severely underestimates values
in early periods and overestimates in recent periods. The average monthly increase
of approximately 13 index points per month is not meaningful given the non-linear
nature of inflation compounding. A log transformation would be more appropriate
but is outside the scope of required methods.

## Method 7: Seasonal Indices

```{r m7-seasonal}
# Seasonal indices measure the relative strength of each month.
# Index > 1: above-average month, Index < 1: below-average month.
# Method: seasonal index = monthly average / overall average

cpi_si      <- cpi_long |> mutate(month_num = month(date))
monthly_avg <- cpi_si |> group_by(month_num) |>
  summarise(avg = mean(index), .groups = "drop")
overall_avg <- mean(actual)

# Compute seasonal indices for all 12 months
seasonal_idx <- monthly_avg$avg / overall_avg
names(seasonal_idx) <- month.abb

cat("Seasonal Indices by Month:
")
print(round(seasonal_idx, 4))
cat("
Highest season:", month.abb[which.max(seasonal_idx)],
    "(index:", round(max(seasonal_idx), 4), ")
")
cat("Lowest season :", month.abb[which.min(seasonal_idx)],
    "(index:", round(min(seasonal_idx), 4), ")
")

# Deseasonalize the series, fit trend, reseasonalize
cpi_si <- cpi_si |>
  left_join(monthly_avg |> rename(month_avg = avg), by = "month_num") |>
  mutate(deseasoned = index / (month_avg / overall_avg))

lm_si      <- lm(deseasoned ~ I(1:n), data = cpi_si)
si_trend   <- fitted(lm_si)
si_fitted  <- si_trend * seasonal_idx[cpi_si$month_num]

# Forecast January 2026 (month 1 seasonal index applied)
si_trend_fc <- coef(lm_si)[1] + coef(lm_si)[2] * (n + 1)
si_forecast <- si_trend_fc * seasonal_idx[1]

cat("Seasonal Index Forecast (January 2026):", round(si_forecast, 2), "
")

acc7 <- calc_accuracy(actual, si_fitted, "7. Seasonal Indices")
accuracy_list[[7]] <- acc7
print(acc7)

plot_forecast(cpi_long, si_fitted, si_forecast,
              forecast_date, "Seasonal Indices", "seasonal_index_plot.png")
```

**Interpretation:** The seasonal indices reveal that January and February tend to
have slightly above-average values while summer months are somewhat below average.
However, MAPE of 117.765% shows poor overall fit because the method uses a linear
trend on deseasonalized data, which fails to capture the exponential growth.
The seasonal component itself is relatively weak compared to the dominant trend.

## Method 8: Additive Decomposition

```{r m8-additive}
# Additive decomposition: Y(t) = Trend + Seasonal + Random
# Assumes seasonal effects are constant in absolute terms regardless of the level.
# Less appropriate here because seasonal variation appears to grow with the series.

decomp_add   <- decompose(cpi_ts, type = "additive")
trend_add    <- as.numeric(decomp_add$trend)
seasonal_add <- as.numeric(decomp_add$seasonal)

# Extend trend using linear regression on valid trend values
vi_add    <- which(!is.na(trend_add))
lm_add    <- lm(trend_add[vi_add] ~ vi_add)
trend_all <- coef(lm_add)[1] + coef(lm_add)[2] * (1:n)
add_fitted   <- trend_all + seasonal_add

# Forecast: extended trend + January seasonal component
add_forecast <- (coef(lm_add)[1] + coef(lm_add)[2] * (n+1)) + seasonal_add[1]

cat("Additive Decomposition Forecast (January 2026):", round(add_forecast, 2), "
")

vi8  <- which(!is.na(add_fitted))
acc8 <- calc_accuracy(actual[vi8], add_fitted[vi8], "8. Additive Decomposition")
accuracy_list[[8]] <- acc8
print(acc8)

plot_forecast(cpi_long, add_fitted, add_forecast,
              forecast_date, "Additive Decomposition", "additive_decomp_plot.png")
```

**Interpretation:** Additive decomposition achieves MAPE of 104.797%, showing poor
fit similar to the linear trend method. The additive model assumes seasonal effects
are constant in absolute magnitude, but in reality the seasonal fluctuations grow
proportionally with the rapidly increasing price level. This violates the core
assumption of additive decomposition, making multiplicative decomposition
theoretically more appropriate for this series.

## Method 9: Multiplicative Decomposition

```{r m9-multiplicative}
# Multiplicative decomposition: Y(t) = Trend * Seasonal * Random
# Assumes seasonal effects are proportional to the trend level.
# More appropriate than additive when seasonal variation grows with the series.

decomp_mul   <- decompose(cpi_ts, type = "multiplicative")
trend_mul    <- as.numeric(decomp_mul$trend)
seasonal_mul <- as.numeric(decomp_mul$seasonal)

# Extend trend using linear regression
vi_mul    <- which(!is.na(trend_mul))
lm_mul    <- lm(trend_mul[vi_mul] ~ vi_mul)
trend_allm <- coef(lm_mul)[1] + coef(lm_mul)[2] * (1:n)
mul_fitted  <- trend_allm * seasonal_mul

# Forecast: extended trend * January seasonal index
mul_forecast <- (coef(lm_mul)[1] + coef(lm_mul)[2] * (n+1)) * seasonal_mul[1]

cat("Multiplicative Decomposition Forecast (January 2026):", round(mul_forecast, 2), "
")

vi9  <- which(!is.na(mul_fitted))
acc9 <- calc_accuracy(actual[vi9], mul_fitted[vi9], "9. Multiplicative Decomposition")
accuracy_list[[9]] <- acc9
print(acc9)

plot_forecast(cpi_long, mul_fitted, mul_forecast,
              forecast_date, "Multiplicative Decomposition", "multiplicative_decomp_plot.png")
```

**Interpretation:** Multiplicative decomposition (MAPE: 104.863%) performs similarly
to additive decomposition despite being theoretically more appropriate. Both methods
fail because they rely on a linear trend extrapolation, which cannot capture the
exponential acceleration observed after 2022. The trend component extracted from
the middle of the series does not represent the recent steep increase adequately.

## Method 10: Regression with Trend and Seasonal Dummy Variables

```{r m10-dummy}
# Multiple regression with time trend and 11 monthly dummy variables.
# Model: Y = a + b*t + sum(gamma_m * D_m) for m=1..11
# December is the reference category (no dummy).
# This method explicitly models both trend and seasonality simultaneously.

# Create 11 monthly dummy variables (January through November)
month_dummies           <- model.matrix(~factor(month(cpi_long$date))-1)[, 1:11]
colnames(month_dummies) <- paste0("M", 1:11)

# Fit the regression model
reg_data     <- data.frame(index = actual, t = 1:n, month_dummies)
lm_dummy     <- lm(index ~ ., data = reg_data)
dummy_fitted <- as.numeric(fitted(lm_dummy))

# Forecast January 2026: t=n+1, M1=1 (January dummy), others=0
new_row           <- data.frame(t = n+1, matrix(0, 1, 11))
colnames(new_row) <- c("t", paste0("M", 1:11))
new_row$M1        <- 1
dummy_forecast    <- as.numeric(predict(lm_dummy, newdata = new_row))

cat("Regression Equation Summary:
")
cat("  Intercept (a)      :", round(coef(lm_dummy)["(Intercept)"], 2), "
")
cat("  Trend slope (b)    :", round(coef(lm_dummy)["t"], 4),
    "- average monthly increase in index points
")
cat("  January effect (M1):", round(coef(lm_dummy)["M1"], 2),
    "- January is above/below December by this amount
")
cat("Dummy Regression Forecast (January 2026):", round(dummy_forecast, 2), "
")

acc10 <- calc_accuracy(actual, dummy_fitted,
                       "10. Trend + Seasonal Dummy Regression")
accuracy_list[[10]] <- acc10
print(acc10)

plot_forecast(cpi_long, dummy_fitted, dummy_forecast,
              forecast_date, "Trend + Seasonal Dummy Regression",
              "regression_seasonal_dummy_plot.png")
```

**Interpretation:** The dummy regression model (MAPE: 118.892%) performs similarly
to linear trend projection because the seasonal dummy variables add little
improvement when the dominant problem is the non-linear trend. The model confirms
that January has a positive seasonal effect relative to December. Like the linear
trend method, this approach fails to capture the exponential growth pattern
of Turkish CPI during the high inflation period.

# Forecast Accuracy Comparison Table

```{r accuracy-table}
# Combine all accuracy results into a single comparison table
accuracy_df <- bind_rows(accuracy_list)

# Add next-period forecast values to the comparison table
forecast_vals <- c(
  naive_forecast, ma_forecast, wma_forecast,
  ses_forecast, holt_forecast, lm_forecast,
  si_forecast, add_forecast, mul_forecast, dummy_forecast
)
accuracy_df$Next_Period_Forecast <- round(forecast_vals, 2)

# Display the full comparison table
print(accuracy_df, digits = 3)

# Save to CSV
write.csv(accuracy_df, "outputs/tables/accuracy_comparison.csv", row.names = FALSE)

# Identify best method by MAPE
best_method <- accuracy_df |> slice_min(MAPE, n = 1)
cat("
=== BEST METHOD BY MAPE ===
")
cat("Method:", best_method$Method, "
")
cat("MAPE  :", round(best_method$MAPE, 3), "%
")
cat("MAD   :", round(best_method$MAD, 3), "
")
```

## Accuracy Comparison Bar Chart

```{r acc-plot}
# Bar chart of MAPE values for all methods
p_acc <- ggplot(accuracy_df, aes(x = reorder(Method, MAPE), y = MAPE, fill = MAPE)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = round(MAPE, 2)), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_gradient(low = "#27AE60", high = "#E74C3C") +
  labs(
    title = "MAPE Comparison Across All Forecasting Methods",
    x     = "Method",
    y     = "MAPE (%)"
  ) +
  theme_minimal(base_size = 11)

print(p_acc)
ggsave("outputs/figures/accuracy_comparison_plot.png", p_acc,
       width = 10, height = 6, dpi = 150)
```

# Superior Method Selection

```{r best-method}
cat("=== SUPERIOR METHOD SELECTION ===

")
cat("Selected Method: Holt Trend-Corrected Exponential Smoothing

")
cat("Quantitative justification:
")
cat("  - Lowest MAPE: 0.825% (vs next best: 1.433% for Naive)
")
cat("  - Lowest MAD : 5.774
")
cat("  - Lowest MSE : 205.945

")
cat("Structural justification:
")
cat("  - Series has STRONG upward trend: Holt explicitly models level + trend
")
cat("  - Alpha and Beta parameters adapt automatically to recent changes
")
cat("  - Unlike MA or SES, Holt does not lag behind during rapid trend changes
")
cat("  - Tracking Signal (39.847) exceeds +/-4 due to the 2022 structural break,
")
cat("    but this is the best among all methods and reflects the data complexity

")
cat("Why other methods are inferior:
")
cat("  - Naive/SES: ignore trend, systematically underestimate
")
cat("  - Moving Average: heavy lag, worst MAPE among single-variable methods
")
cat("  - Linear/Dummy/Seasonal: assume linear trend, fail on exponential growth
")
cat("  - Decomposition: linear trend extrapolation fails post-2022
")
```

## Superior Method Plot

```{r best-plot}
# Final plot: actual series + Holt fitted + January 2026 forecast
superior_plot <- ggplot(cpi_long, aes(x = date, y = index)) +
  geom_line(color = "#2C3E50", linewidth = 0.8) +
  geom_line(aes(y = holt_fitted), color = "#E74C3C",
            linewidth = 0.7, linetype = "dashed") +
  geom_point(data = data.frame(date = forecast_date, index = holt_forecast),
             aes(x = date, y = index), color = "#27AE60", size = 5, shape = 18) +
  annotate("text", x = forecast_date, y = holt_forecast + 80,
           label = paste("Jan 2026
", round(holt_forecast, 2)),
           color = "#27AE60", size = 3.5, fontface = "bold") +
  labs(
    title    = "Superior Method: Holt Trend-Corrected Exponential Smoothing",
    subtitle = paste("January 2026 Forecast:", round(holt_forecast, 2),
                     "(2003=100 base)"),
    x        = "Date",
    y        = "CPI Index (2003=100)",
    caption  = "Black: Actual | Red dashed: Holt Fitted | Green: January 2026 Forecast"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

print(superior_plot)
ggsave("outputs/figures/superior_method_plot.png", superior_plot,
       width = 10, height = 5, dpi = 150)
```

# Final Forecast

```{r final-forecast}
# Compile all forecasts into a summary table
all_forecasts <- data.frame(
  Method           = accuracy_df$Method,
  Forecast_Jan2026 = round(forecast_vals, 2),
  MAPE             = accuracy_df$MAPE
) |> arrange(MAPE)

print(all_forecasts)

# Final result
cat("
================================================
")
cat("FINAL FORECAST RESULT
")
cat("================================================
")
cat("Superior Method    : Holt Trend-Corrected ES
")
cat("Date of Access     :", as.character(Sys.Date()), "
")
cat("Latest Observation : December 2025 (Index:", round(tail(actual,1),2), ")
")
cat("Forecast Period    : January 2026
")
cat("Forecasted Value   :", round(holt_forecast, 2), "(2003=100 base)
")
cat("================================================
")

# Save final forecast CSV with all required fields
final_df <- data.frame(
  Superior_Method        = "Holt Trend-Corrected Exponential Smoothing",
  Date_of_Data_Access    = as.character(Sys.Date()),
  Latest_Observation     = "December 2025",
  Latest_Index_Value     = round(tail(actual, 1), 2),
  Forecast_Target_Period = "January 2026",
  Forecasted_Value       = round(holt_forecast, 2),
  Base_Year              = "2003=100",
  MAPE                   = acc5$MAPE,
  MAD                    = acc5$MAD,
  Bias_ME                = acc5$Bias_ME,
  MSE                    = acc5$MSE,
  Tracking_Signal        = acc5$Tracking_Signal
)
write.csv(final_df, "outputs/tables/final_forecast.csv", row.names = FALSE)
cat("final_forecast.csv saved.
")
```

# Interpretation of Results

The Turkish CPI General Index is forecast to reach approximately **3570.68** in
January 2026, up from 3513.87 in December 2025. This represents a monthly increase
of approximately **1.6%**, which is consistent with the gradual deceleration in
Turkish inflation observed throughout 2025 (annual inflation fell from over 75% in
early 2024 to approximately 40% by late 2025).

The Holt method captures this decelerating trend by continuously updating both
the level and trend estimates. The positive but moderating beta parameter reflects
the slowing pace of price increases. The forecast is reasonable given current
economic conditions in Turkey.

# Limitations

1. **API Authentication:** TÜİK SDMX API returns HTTP 401; data accessed via
   alternative R-based method (instructor-approved).
2. **Structural Break:** The 2022 currency crisis created a non-linear acceleration
   that classical methods cannot fully model.
3. **Classical Methods Only:** ARIMA, ETS, and machine learning models are excluded
   as per assignment requirements.
4. **Tracking Signal:** Most methods show Tracking Signal > ±4 due to the 2022-2023
   high inflation period; this does not indicate model failure but reflects the
   extraordinary nature of that period.
5. **Point Forecast Only:** No confidence intervals are reported.
6. **External Shocks:** Geopolitical events, policy changes, or exchange rate
   movements are not incorporated into the forecast.

# Reproducibility Notes

To reproduce this analysis:

1. Install R (version >= 4.2) and RStudio
2. Run the following in R:

```r
install.packages(c("forecast","zoo","ggplot2","tidyr",
                   "lubridate","readxl","httr","rmarkdown","devtools"))
devtools::install_github("emraher/tuikr")
setwd("path/to/tuik-forecasting-project")
rmarkdown::render("forecasting_project.Rmd")
```

Alternatively, restore the exact package environment:
```r
renv::restore()
rmarkdown::render("forecasting_project.Rmd")
```

Data is fetched automatically from the TÜİK server on each run.
No local data files are required.

